
Storm consequences across the United States.
==================================

#Synopsis
The aim of this report is to understand the type of storm events accros  
the Untied States that are most harmful for the populations healt and have  
the greatest econommic consequences. 



#Loading and processing the raw data
 

The data is uploaded form the course web site [Storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)  
  
Documentation describing the database is available at the following locations 
- National Wather service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

## Reading the data 
First read the data form the bz2 commpressed file, the fields are delimited by a ,  
To avoid the long reading time of the file every time the block cache option is enabled. 
```{r loaddata, cache=TRUE}
stormdata <- read.table("repdata_data_StormData.csv.bz2", sep = ",",na.strings = "")
```

After reading we test the first rows, there should be 902298 observations and 37 vaiables.
```{r}
dim(stormdata)
```
```{r}
head(stormdata[, 1:10])
```
